---
title: "Estimating sample size for correlation analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating sample size for correlation analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(qtils)
```

Here we're providing some basic dogfooding for performing sample size estimation for correlation-based analysis, especially to leverage simulations.   

## Intraclass correlation analysis for test-retest reliability. 

Often we are required to perform sample size analysis for test-retest reliability, which measures the differences between two consecutive measures of the same 



## References. 

- Stats exchange [topic](https://web.archive.org/web/20240111193757/https://stats.stackexchange.com/questions/263451/create-synthetic-data-with-a-given-intraclass-correlation-coefficient-icc) regarding simulating with specific ICC(1,1) in mind 



